---
#title: "Mouse and rat transcript analysis"
#author: "gwct"
#date: "04/14/2020"
output: 
  html_document:
    includes:
      in_header: '../html-chunks/rmd_header.html'
    df_print: paged
    code_folding: hide
---

```{r setup}
binData <- function(df, column, max_bin){
# Function to pre-bin SVs by length
  binned_df = data.frame("bin"=seq(from=0,to=max_bin,by=0.01), count=0)
  for(i in 1:nrow(df)) {
    row <- df[i,]
    bin = floor(row[[column]]*100)/100
    binned_df[binned_df$bin==bin,]$count = binned_df[binned_df$bin==bin,]$count + 1
  }
  return(binned_df) 
}

knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(plyr)
source("../lib/design.r")

htmltools::includeHTML("../html-chunks/nav.html")
```

# Exome transcript selection

## Filtering criteria:

1. Transcript must be present in both mouse and rat.
2. Transcript must have one-to-one relationship between mouse and rat.
3. Orthology confidence must be high (1).
4. Transcript must have dS below 0.5 (see distibutions below).
5. If multiple transcripts from a gene pass all above filters, the one with the average longest length between mouse and rat is selected.

## Filter script log

```
Rodent exomes -- select mouse trancsripts
PYTHON VERSION: 3.6.6
Script call: select_transcripts.py
Runtime: 04/14/2020 21:09:10
----------
Mouse GTF file:        ../02-Annotation-data/reference-data/Mus_musculus.GRCm38.99.gtf.gz
Rat GTF file:          ../02-Annotation-data/reference-data/Rattus_norvegicus.Rnor_6.0.99.gtf.gz
Ensembl ortholog file: ../02-Annotation-data/reference-data/mouse-rat-orths-ens99.txt
Output file:           ../02-Annotation-data/selected-transcripts.txt
--------------
dS threshold:          0.5
--------------
04.14.2020 | 21:09:10 Reading mouse feature lengths...
Mouse transcripts read: 142647
04.14.2020 | 21:09:20 Reading rat feature lengths...
Rat transcripts read:   41078
--------------
04.14.2020 | 21:09:27 Reading orthologs...
Genes read:      56289
Transcipts read: 170971
--------------
04.14.2020 | 21:09:28 Filtering transcripts...
Transcripts with no rat ortholog:                                                      44696
Transcripts with no one2one rat ortholog:                                              37291
Transcripts with no high confidence rat ortholog:                                      11069
Transcripts with no rat ortholog below dS threshold:                                   1204
Transcripts filtered for not being, on average between rat and mouse, longest in gene: 56413
Note: numbers reported are conditional on those trancsripts passing the previous filter. Filters reported in order of their coding.
--------------
Genes with no passed transcripts:                                                      41857
Genes with selected transcript:                                                        14432
--------------
04.14.2020 | 21:09:28 Writing selected transcripts to output file...
```

```{r readdata, echo=FALSE, warning=FALSE}
pre_data = read.csv("../../../02-Annotation-data/reference-data/mouse-rat-orths-ens99.txt", header=T, sep="\t")
pre_data$dnds = pre_data$dN.with.Rat / pre_data$dS.with.Rat
pre_data = subset(pre_data, dS.with.Rat < 1)
pre_data = subset(pre_data, dN.with.Rat < 1)


filter_data = read.csv("../../../02-Annotation-data/selected-transcripts.txt", header=T, sep="\t", comment.char='#')
filter_data$dnds = filter_data$dN.with.Rat / filter_data$dS.with.Rat
filter_data = subset(filter_data, dS.with.Rat < 1)
filter_data = subset(filter_data, dN.with.Rat < 1)
```

## Mouse-rat dS for orthologous transcripts -- All transcripts

Data from Ensembl release 99.

```{r fig1, out.width="100%", fig.align = "center", warning=FALSE}
#####################
# Mouse-rat dS for transcripts
#####################\
pre_binned = binData(pre_data, "dS.with.Rat", 1)
pre_binned$filter = "Pre-filter"

filter_binned = binData(filter_data, "dS.with.Rat", 1)
filter_binned$filter = "Post-filter"

binned_data = rbind.fill(pre_binned, filter_binned)
col = corecol(numcol=2)
lab = c("Pre-filter", "Post-filter")
ylabels = c("1200", as.character(seq(from=0, to=7000, by=1000)))

dualp = ggplot(binned_data, aes(x = bin, y = ifelse(filter == "Post-filter",-1, 1)*count, fill = filter)) + 
  geom_col() +
  geom_hline(yintercept=0) +
  scale_x_continuous(name = "dS", limits = c(-0.02, 1), expand = c(0, 0)) +
  scale_y_continuous(name = "# Transcripts", limits=c(-1200,7000), breaks = 1000*(-1:7), labels=ylabels) +
  scale_fill_manual(name="", limits=lab, values=col) +
  bartheme()
print(dualp)
```

## Mouse-rat dN for orthologous transcripts

Data from Ensembl release 99.

```{r fig2, out.width="100%", fig.align = "center", warning=FALSE}
#####################
# Mouse-rat dN for transcripts
#####################\
pre_binned = binData(pre_data, "dN.with.Rat", 1)
pre_binned$filter = "Pre-filter"

filter_binned = binData(filter_data, "dN.with.Rat", 1)
filter_binned$filter = "Post-filter"

binned_data = rbind.fill(pre_binned, filter_binned)
col = corecol(numcol=2)
lab = c("Pre-filter", "Post-filter")
ylabels = c("5000", as.character(seq(from=0, to=30000, by=5000)))

dualp = ggplot(binned_data, aes(x = bin, y = ifelse(filter == "Post-filter",-1, 1)*count, fill = filter)) + 
  geom_col() +
  geom_hline(yintercept=0) +
  scale_x_continuous(name = "dN", limits = c(-0.02, 1), expand = c(0, 0)) +
  scale_y_continuous(name = "# Transcripts", limits=c(-5000,30000), breaks = 5000*(-1:6), labels=ylabels) +
  scale_fill_manual(name="", limits=lab, values=col) +
  bartheme()
print(dualp)
```

## Mouse-rat dN/dS for orthologous transcripts

Data from Ensembl release 99.

```{r fig3, out.width="100%", fig.align = "center", warning=FALSE}
#####################
# Mouse-rat dN/dS for transcripts
#####################
pre_data = subset(pre_data, !is.na(dnds))
pre_binned = binData(pre_data, "dnds", 2)
pre_binned$filter = "Pre-filter"

filter_data = subset(filter_data, !is.na(dnds))
filter_binned = binData(filter_data, "dnds", 2)
filter_binned$filter = "Post-filter"

binned_data = rbind.fill(pre_binned, filter_binned)
col = corecol(numcol=2)
lab = c("Pre-filter", "Post-filter")
ylabels = c("1000", as.character(seq(from=0, to=6000, by=1000)))

dualp = ggplot(binned_data, aes(x = bin, y = ifelse(filter == "Post-filter",-1, 1)*count, fill = filter)) + 
  geom_col() +
  geom_hline(yintercept=0) +
  scale_x_continuous(name = "dN/dS", limits = c(-0.02, 2), expand = c(0, 0)) +
  scale_y_continuous(name = "# Transcripts", limits=c(-1000,6000), breaks = 1000*(-1:6), labels=ylabels) +
  scale_fill_manual(name="", limits=lab, values=col) +
  bartheme()
print(dualp)
```



```{r footer}
cat("Page last updated:", format(Sys.time(), "%m/%d/%Y %H:%M:%S %Z"))
htmltools::includeHTML("../html-chunks/rmd_footer.html")
```